Plant Modeling Using Environmental Parameters

### 2014-03-28 doc/paper.tex
Webpage, abstract

@@

The project will focus on modeling natural looking plants using simulated
environment parameters to control growth and evolution of different plant
species. Genetic algorithms or other evolutionary techniques will be used to
evolve plant suitable for growth in various environments.

Potential environmental conditions:

\begin{itemize*}
  \item Availability of water
  \item Amount of sunlight
  \item Soil conditions
  \item Growth elevation
\end{itemize*}

Potential plant species parameters:

\begin{itemize*}
  \item Rate of water loss
  \item Food and water storage capability
  \item Total mass/volume
  \item Crown height
  \item Leaf area
  \item Root structures
\end{itemize*}

@@

\section{Introduction}

Karl Sims introduced Panspermia in 1990.

LSystem

We closely examined different existing frameworks that we considered using for
our project. The two frameworks that were the best candidates was
OpenAlea\cite{openalea} and Algorithmic Botany\cite{abotany}.

@@

The software was implemented using Java and OpenGL for drawing. The top-level
architecture can be divided into two main categories: plant simulation using
L-Systems and the graphical user interface.

@@

The L-Systems are implemented in the LSystem and LRule classes, which are
responsible for storing the production rules associated with each plant and
iterating the plants phenotype as the plant grows throughout the simulation. The
Gene and Plant classes store representations of each living plant. The Gene
class represents the plant genotype as a combination of the L-System production
rules along with additional information such as stem and leaf size, stem and
flower colors, the number of seeds produced, and the age at which the plant
matures. Each individual plants phenotype is represented using the Plant class
and contains it's X/Y position and orientation inside the world, along with the
current age and a L-System string (L-String) representing the plants current
shape. The Plant class is also responsible for calculating each individual
plants fitness values at the end of each simulation tick.

@@

\includegraphics[width=\columnwidth]{images/architecture.png}

@@

Each plant in our environment contains a string of symbol (which we call an
"L-String") representing the plants current shape and size. Each symbol can
represent a stem segment, a flower, a leaf, etc. The plants genotype contains a
set of production rules that govern the growth and development of the plant.
When the plant grows, the production rules are used to replace symbols in the
plants current L-String with sequences of tokens to produces a larger and more
complex plant.

\subsection{L-System Genotypes}

We implemented a modified variant of L-Systems where a replacement probability
is included in addition to the matching and replacement parts of each production
rule. The possible values for each symbol in our rules are as follows:

@@

\begin{description}
  \item[Stem color]   \hfill \\
    The plants stem color is changed randomly, the stem color color is also used
    for the plants leaves.
  \item[Flower color] \hfill \\
    The plants flower color is changed randomly. Both the stem color and flower
    color are only used for display and can be used to help differentiate plants
    of different species.
  \item[Stem width]   \hfill \\
    The diameter of the plants stem is changed. Upper and lower limits are
    imposed and a width is randomly chosen from within the range.
  \item[Leaf size]    \hfill \\
    The length of each leaf is changed. Upper and lower limits are used. The
    leaf shape is fixed for drawing purposes.
  \item[Growth rules] \hfill \\
    The L-System productions used to control the plants growth are changed. The
    plant may become either more or less complex.
\end{description}

@@

Overall, we have shown that artificial plants based on L-Systems can be evolved
in realistic ways based on abstract environmental parameters. Using L-Strings to
represent plants results in both realistic looking shapes and provides the
foundation for genetic algorithm for controlling the development of plant
species.

>>>>

\newcommand{\onecol}[3]{
  \begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{#1}
    \vspace{0.0in}%??
    \caption{#3}
    \label{#2}
  \end{figure}
}

\newcommand{\twocol}[3]{
  \begin{figure*}
    \centering
    \includegraphics[width=5.5in]{#1}
    \vspace{0.1in}%??
    \caption{#3}
    \label{#2}
  \end{figure*}
}

@@

L-systems, or Lindenmayer systems, are often used for realistic looking
artificial plant environments. While conceptually simple, L-systems have been
extended with many features to create realistic plants models based on a
description of the plant growth patterns. While significant progress has been
made in providing robust frameworks for modeling artificial plants, the process
of creating specific plants species is still very much a manual process that
often requires a deep understanding of the tools used.

In this project we will explore an alternate approach to plant species creation
by using genetic algorithms to automatically evolve plants suitable to
particular environments.

By tuning environmental parameters for availability of water, the amount of
sunlight, and the ambient soil conditions, artificial plant species are produced
which resembling physical plants that grown in similar environments.

In this approach to plant generation, the underlying L-system production rules
are evolved based on random mutation combined with a fitness function to
determine plant survival rates. In addition to the L-systems, several additional
parameters are introduced to control plant growth and simulation at a more
abstract level.

@@

@@

We closely examined different existing frameworks that we considered using for
our project. The two frameworks that were the best candidates was
OpenAlea\cite{openalea} and Algorithmic Botany\cite{abotany}. Eventually we
decided to implement the software from scratch.

The software was implemented using Java and OpenGL for drawing. The top-level
architecture can be divided into two main categories: plant simulation using
L-systems and the graphical user interface.

@@

The L-systems are implemented in the LSystem and LRule classes, which are
responsible for storing the production rules associated with each plant and
iterating the plants phenotype as the plant grows throughout the simulation. The
Gene and Plant classes store representations of each living plant. The Gene
class represents the plant genotype as a combination of the L-system production
rules along with additional information such as stem and leaf size, stem and
flower colors, the number of seeds produced, and the age at which the plant
matures. Each individual plants phenotype is represented using the Plant class
and contains it's X/Y position and orientation inside the world, along with the
current age and a L-system string (L-string) representing the plants current
shape. The Plant class is also responsible for calculating each individual
plants fitness values at the end of each simulation tick.

@@

\onecol{images/architecture.png}{arch}{Conceptual software architecture}

@@

Each plant in our environment contains a string of symbol (which we call an
"L-string") representing the plants current shape and size. Each symbol can
represent a stem segment, a flower, a leaf, etc. The plants genotype contains a
set of production rules that govern the growth and development of the plant.
When the plant grows, the production rules are used to replace symbols in the
plants current L-string with sequences of tokens to produces a larger and more
complex plant.

\onecol{images/screenshot.png}{screen}{Example plant simulation}

\subsection{L-system Genotypes}

We implemented a modified variant of L-systems where a replacement probability
is included in addition to the matching and replacement parts of each production
rule. The possible values for each symbol in our rules are as follows:

@@

\begin{description}
  \item[Stem color]   \hfill \\
    The plants stem color is changed randomly, the stem color color is also used
    for the plants leaves.
  \item[Flower color] \hfill \\
    The plants flower color is changed randomly. Both the stem color and flower
    color are only used for display and can be used to help differentiate plants
    of different species.
  \item[Stem width]   \hfill \\
    The diameter of the plants stem is changed. Upper and lower limits are
    imposed and a width is randomly chosen from within the range.
  \item[Leaf size]    \hfill \\
    The length of each leaf is changed. Upper and lower limits are used. The
    leaf shape is fixed for drawing purposes.
  \item[Growth rules] \hfill \\
    The L-system productions used to control the plants growth are changed. The
    plant may become either more or less complex.
\end{description}

@@

Overall, we have shown that artificial plants based on L-systems can be evolved
in realistic ways based on abstract environmental parameters. Using L-strings to
represent plants results in both realistic looking shapes and provides the
foundation for genetic algorithm for controlling the development of plant
species.
